What We Eval in the Shadows

### 2021-04-16 paper/main.tex
Conclusion: advocating for macros in R

% \section{\eval expressivity and macros}

% The Thomas Lumley's article, Programmer's Niche: Macros in R, (https://www.r-project.org/doc/Rnews/Rnews_2001-3.pdf) describes how to write a `defmacro` function that would like as a macro creator, using `substitute` and `eval`.  There are no local macro variables in this implementation though.

@@

% how much can we say about the threat model?
Similarly to \citep{ecoop11}, we started looking into \evals in R with the
ambition to replace them by other feature that are more amenable to static
analysis. The data unfortunately suggests that replacing \eval is going to be
a lengthy journey. Even if we disregard core packages as they are part of the
language and the \eval use is stable, in CRAN \eval is used widely and in more
sophisticated way than was the case in JavaScript. There the vast majority of
sites fitted an extremely simple categorization based on simple regular
expressions. Not so in R. While some usage, especially the one which calls
\eval with an expression parsed from string, can be categorized and often
replaced by equivalent yet more disciplined and safer code, the case of \eval
coming from reflection is much harder.

>>>>

@@

% how much can we say about the threat model?
Similarly to \citep{ecoop11}, we started looking into \evals in R with the
ambition to replace them by other features that are more amenable to static
analysis. The data unfortunately suggests that replacing \eval is going to be
a lengthy journey. Even if we disregard core packages as they are part of the
language and the \eval use is stable, in CRAN \eval is used widely and in more
sophisticated way than was the case in JavaScript. There the vast majority of
sites fitted an extremely simple categorization based on simple regular
expressions. Not so in R. While some usage, especially the one which calls
\eval with an expression parsed from string, can be categorized and often
replaced by equivalent yet more disciplined and safer code, the case of \eval
coming from reflection is much harder.

Even though we reckon that \eval has more expressive power than macros, as it can access runtime information such as its call stack, adding a macro system to R to replace \eval for all meta-programming purposes would help static analysis as macros can be expanded at analysis-time.

% It's dubious whether R programmers would develop with macros though.
% Macros and R
% The Thomas Lumley's article, Programmer's Niche: Macros in R, (https://www.r-project.org/doc/Rnews/Rnews_2001-3.pdf) describes how to write a `defmacro` function that would like as a macro creator, using `substitute` and `eval`.  There are no local macro variables in this implementation though.
